A significance test for forward stepwise model selection

### 2014-05-12 paper/paper.tex
Paper edits, adding figs/tables

% "Title of the paper"
\title{A significance test in forward stepwise model selection}

@@

\begin{abstract}
% perhaps de-emphasize group LASSO and emphasize forward stepwise?
  We extend the methods developed by \cite{significance:lasso} and
  \cite{tests:adaptive} on significance tests for penalized
  regression to stepwise model selection by iteratively applying
  the global null hypothesis test on the residual in a forward stepwise
  procedure. While not fully theoretically justified, the resulting method
  has the computational strengths of stepwise selection and partially
  solves the problem of inflated test statistics due to model selection.
  We illustrate the flexibility of this method by applying it to
  novel specialized applications of forward stepwise to a hierarchically
  constrained interactions model and generalized additive models.
\end{abstract}

@@

Consider the regression setting with a single response variable $Y$
and collection of predictor or covariate variables denoted $\mathcal X$.
One often wishes to choose a subset of variables $X \subset \mathcal X$
for modeling the response, assuming the remaining variables are
irrelevent and can be discarded without much loss of predictive
or explanatory power. Doing this in a structured and principled way
usually requires algorithmic methods for choosing $X$. One such method,
forward stepwise regression, is a procedure
that begins with an empty model and sequentially adds the best predictor
variable in each step. Because of the stochastic nature of this
algorithm--making use of the data to choose variables--
the usual $\chi^2$ and $F$-tests for significance
fail when a model has been selected this way and tend to be
anti-conservative unless they are computed on a held-out validation
dataset. This is one instance of the general problem of conducting
inference and model selection using the same data, a problem
of central importance on which some recent progress has been made.

In the LASSO setting, \cite{significance:lasso} derived
a novel test statistic and its asymptotic null distribution, making
possible valid inferences after model selection using the full data.
\cite{tests:adaptive} modified and extended those results to the
group LASSO \citep{grouplasso} and other penalized regression
problems, but only under the global null hypothesis.
One of the strengths of these test statistics is that they can be
used for valid significance testing when computed on the same
data used for model selection, eliminating the need for data splitting
or for situations in which data splitting is not appropriate. For example,
when categorical covariates contain rare levels with very few observations
it can be difficult or impossible to split the sample with an adequate
number of such observations occurring in each split.

@@

As a small illustrative example, consider a response $Y$ with ten
categorical predictors $X_g$ having between 2 and 4 levels. The true 
relationship is that $Y$ only depends on $X_1$ and $X_9$. Specifically,
if observation $i$ has covariates $X_{1,i} = j, X_{9,i} = k$ then
\[
    Y_i = \beta_{1,j} + \beta_{9,k} + \epsilon_i , \qquad
    \epsilon_i \sim N(0,1)
\]
$X_1$ has three levels with $\beta_1 = (1, .5, -1)$, $X_9$
has two levels with $\beta_9 = (.5, -.5)$. We generated a random
categorical design matrix $X$ and created an instance of this example
with $n = 40$ observations. Running forward stepwise, we calculated
both our p-value and the usual $\chi^2$ p-value at each step. In
Table~\ref{tab:example} we see that forward stepwise chooses the
truly nonzero variables first and both p-values for these are small.
However, once forward stepwise begins adding noise variables, the
$\chi^2$ p-value remains relatively small, potentially leading to 
incorrect inferences.

@@

In the next section we establish general notation and describe the
forward stepwise algorithm used throughout the paper.
Section~\ref{sec:testing} reviews some recent work on
post-selection inference
\citep{significance:lasso,tests:adaptive,lasso:fixed}
relevant to our work here, and describes a general framework
for post-selection inference based on quadratic comparisons.
Simulation results in Section~\ref{sec:simulations} show
empirically that our method performs well in settings where forward
stepwise itself performs well, and that
various stopping rules using our test statistic--including
some from \cite{sequential:fdr}--appear promising. In
Section~\ref{sec:applications} we apply
the method to several variants of forward
stepwise tailored to models with interactions and generalized additive
models, as well as to a real data example involving genomic prediction
of individual drug responses to treat HIV.

@@

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of dummy variable encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single functional pathway.
To emphasize this we use $g, h$ as covariate indices rather than the usual
$i, j$ throughout. Since single variables can be considered groups of
size 1, this includes non-grouped situations as a special
case. Our variant of forward stepwise uses a different method than
usual for choosing which group to add because we are using the
same objective in the forward stepwise procedure that was used to
derive the test statistics in penalized regression settings.

Denote $y \in \real^n$ for $n$ i.i.d. measurements of the outcome variable.
Let an integer $G \geq 2$ be the number of groups of explanatory variables.
For each $1 \leq g \leq G$ the design matrix encoding the
$g$th group is the $n \times p_g$ matrix denoted $X_g$, where $p_g$ is
the number of individual variables or columns in group $g$.
% Do we need to assume $p_g \ll n$ or $\text{rank}(X_g) = p_g$?
When a group encodes a categorical variable as indicators for the
levels of that variable, by default we use the full encoding with a
column for every level. Although this introduces collinearity, our method
does not require the design matrix to have full rank.

Denote $p = \sum_{g=1}^Gp_g$,
the total number of individual variables, so $p = G$ in the
case where all groups have size 1. Let $X$ be the matrix constructed
by column-binding the $X_g$, that is  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}
X = \begin{pmatrix} X_1 & X_2 & \cdots & X_G  \end{pmatrix}
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We normalize each group by its Frobenius norm, but this is not required
especially since we also allow weights, $w_g$, for each group. These
weights act like penalties or costs, so increasing $w_g$ makes it
more difficult for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily, but we set them all to be
1 since we already normalize by the Frobenius norm.

With each group we associate the $p_g \times 1$ coefficient vector
$\beta_g$, and write $\beta$ for the $p \times 1$ vector constructed
by stacking all of the $\beta_g$ in order.  Finally, our model for the
response is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gmodel}
y | X& = X \beta + \sigma \epsilon \\
   & = \sum_{g=1}^G X_g \beta_g + \sigma \epsilon
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\epsilon$ is noise. We assume Gaussian noise
$\epsilon | X \sim N(0, \Sigma)$ with known covariance matrix $\Sigma$.

@@

\subsection{Description of the algorithm}

First, the user must specify the maximum number of steps allowed,
which we denote \textit{steps} (consistent with the \textit{R} function
for stepwise). To enable our test statistic computations,
\textit{steps} should be at most $\min (n, G) - 1$, but it is
computationally desirable to set it as low as possible while
safely larger than the sparsity of $\beta$. 
Then forward stepwise may recover all the nonzero
coefficients of $\beta$ and terminate without performing much
additional computation.
Of course the sparsity is usually unknown, so this requires guesswork.
In our implementation we treat the active set $A$ as an ordered list to easily track the order of groups entering the model.

@@

To describe the previous work we require some facts about the solution path
of the LASSO.
Writing $\hat \beta(\lambda)$ for the LASSO solution for a fixed value of $\lambda$, we recall
the following facts summarized in \cite{significance:lasso,tibshirani_LASSO_uniqueness}.

@@

The covariance test (assuming groups of size one and $\Sigma = \sigma^2 I$ is given at the first step by
\begin{equation}
\label{eq:covtest}
T_1 = \frac{\lambda_1(\lambda_1-\lambda_2)}{\sigma^2} \overset{n,p \to \infty}{\to} \text{Exp}(1).
\end{equation}
In \cite{tests:adaptive} it is pointed out 
\begin{equation}
\label{test:exact}
\exp\left(- \frac{\lambda_1(\lambda_1-\lambda_2)}{\sigma^2} \right) \approx \frac{1 - \Phi(\lambda_1/\sigma)}{1 - \Phi(\lambda_2 / \sigma)} \overset{D}{=} \text{Unif}(0,1).
\end{equation}
Further, the above test statistic can be understood as the survival function of $\lambda_1 = \|X^Ty\|_{\infty}$
{\em conditional} on which variable achieves $\lambda_1$ as well as its sign.
The limiting results about later steps in \cite{significance:lasso} can
%(formally)
be interpreted as
recursively applying $T_1$ to the variables that had not previously been selected by LARS.

In the setting of grouped variables, we use an analogous test statistic to
the right hand side of \eqref{test:exact}. The test statistic is presented as an example
in \cite{tests:adaptive} though we re-derive it here in simpler form. The test draws inspiration
from the group LASSO.

Recall the LASSO estimator is given by
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:lasso}
\displaystyle \hat \beta_\lambda = \argmin_{\beta \in \real^p} \frac{1}{2} \| y - X \beta \|_2^2 +
   \lambda \| \beta \|_1 \\
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The \emph{group LASSO estimator}, a generalization of the LASSO,
 is a solution to the following penalized
least-squares convex problem
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gsoln}
\displaystyle \hat \beta_\lambda = \argmin_{\beta \in \real^p} \frac{1}{2} \| y - X \beta \|_2^2 +
   \lambda {\cal P}(\beta) \\
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
with the group penalty
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
  \begin{aligned}
  \label{eq:gpen}
    {\cal P}(\beta)&= \sum_{g=1}^G w_g \| \beta_g \|_2 \\
  \end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The parameter $\lambda \geq 0$ enforces sparsity in groups: for large
$\lambda$ most of the $\beta_g$ will be zero vectors. The weights
$w_g$ are usually taken to be $\sqrt {p_g}$ to normalize the penalty
across groups with different sizes.  This can also be accomplished
by scaling the corresponding submatrices by their Frobenius norms
and setting all weights equal.
Note that this includes the usual LASSO estimator as a
special case when all groups are of size 1, since then the
penalty term is the $L^1$-norm of $\beta$. The solution path of the group
LASSO has similar properties to the LASSO case, however it is not
generally piecewise linear.

@@

With non-trivial groups, however, the event that $g^*(y)$, the first
group chosen, is equal to some fixed group $g$ is generally not given by
a set of affine constraints. Rather, considering line 3 of Algorithm~\ref{algo:fs}, we see
\begin{equation}
\label{eq:first:quadratic}
\begin{aligned}
\{g^*(y)=g\} &= \{ \|X_g^Ty\|_2 /w_g \geq \|X_h^Ty\|_2 /w_h , \ \forall h \neq g\} \\
&=  \{ y^T(X_g^TX_g)y / w_g^2 - y^T(X_h^TX_h)y / w_h^2 \geq 0, \ \forall h \neq g\} \\
\end{aligned}
\end{equation}
Hence, our selection event can be expressed as 
the intersection of a list of quadratic inequalities. This non-affine selection event is stylized in Figure
\ref{fig:curved}.
In the rest of this section, we consider an arbitrary selection procedure
$S$ that returns one of a set of possible outcomes $s \in {\cal S}$ by a set of quadratic inequalities.
That is, for all possible outcomes $s \in {\cal S}$ there is a set of indices
$I(s)$ such that
\begin{equation}
\label{eq:selection}
\begin{aligned}
\{S(y)=s\} = \cap_{i \in I(s)} \{y: y^TQ_iy + a_i^Ty\leq b_i \}.
\end{aligned}
\end{equation}
The quadratic forms are not assumed non-negative definite, but we can, without loss of generality
assume they are symmetric.
As in \cite{lasso:fixed} we will be interested in some randomly chosen contrast
$\eta(S(y))^T\mu$.
In our grouped selection procedure, the selection rule is $S(y)=g^*(y)$
and $\eta$ is typically 
$$
\eta(S(y)) = \eta(g^*(y)) = \frac{X_{g^*(y)}^Ty}{\|X_{g^*(y)}^Ty\|_2}.
$$

Having fixed $\eta$, we proceed to study this contrast by slicing through the selection event along a ray with direction $\eta$ that passes through $y$. 
That is, we need to find
$$
\begin{aligned}
\lefteqn{
\left\{t \in \mathbb R : S(y+t \cdot \eta)=s \right\} } \\
 & \qquad = \cap_{i \in I(s)} \left\{t \in \mathbb R : (y+t \cdot \eta)^T Q_i (y+t \cdot \eta) + a_i^T(y+t \cdot \eta) \leq b_i\right\}. \\
\end{aligned}
$$

\begin{figure}
\label{fig:curved}
\begin{center}
\resizebox{!}{3in}{\input{../figs/curved.pdf_edit_tex}}
\end{center}
A generic selection event, given by the intersection of a list
of quadratic inequalities. Note that the event need not be simply connected. \end{figure}

The slice for any particular inequality is
$$
\begin{aligned}
\lefteqn{
\left\{t \in \mathbb R : (y+t\cdot \eta)^T Q_i (y+t \cdot \eta) + a_i^T(y+t \cdot \eta) \leq b_i\right\}} \\
 & \qquad = \left\{t \in \mathbb R : t^2 \cdot (\eta^TQ_i\eta) + t \cdot(2 y^TQ_i\eta + a_i^T\eta) + y^TQ_iy + a_i^Ty - b_i \leq 0 \right\} \\
\end{aligned}
$$
This can be computed explicitly, and results in either the empty set, a single interval (possibly infinite) or the union of 
two disjoint infinite intervals. The intersection over all $I(s)$ is therefore also computable, yielding a form for the slice
\begin{equation}
\label{eq:slice}
\{g^*(y)=g, \eta(g^*(y))=\eta\}.
\end{equation}
In the picture \ref{fig:curved}, for every $\eta$, the slice \eqref{eq:slice} is a function
of $P_{\eta}^{\perp}y$. This amounts to a proof of the following lemma.

@@

{\bf Proof:}
For any fixed $s$, decompose $y$ as
$$
(Z^{\perp}_s,Z_s) = (P_s^{\perp}y, X_s^Ty).
$$
Under $H_{0,s}:X_s^T\mu=0$, the density of $Z_s$ can be written as
$$
(2 \pi \Sigma_s \sigma^2)^{-k(s)/2} \exp \left(-\frac{1}{2 \sigma^2} z^T\Sigma_s^{-1}z \right)
$$
with $\Sigma_s = X_s^TX_s$.
Transforming to polar coordinates $(R,U)$ yields a density
$$
(2 \pi \Sigma_s \sigma^2)^{-k(s)/2} r^{k(s)-1} \exp \left(-\frac{r^2}{2 \sigma^2} u^T\Sigma_s^{-1}u \right), \qquad r \geq 0, \|u\|_2=1.
$$
Conditioning on $U$ shows that $R=\|X_s^Ty\|_2|U$ has distribution proportional to a $\chi_{k(s)}$. Finally, observe that 
the scale of the $\chi$ above is given by $\theta_s$. \qed
%$\sigma \|X_s\eta_s\|_2$.

@@

\subsection{Computation of p-value}

Lemma \ref{eq:test:dbn} gives an explicit characterization of our p-value
that can also be used to calculate it in practice. However, in this
section we describe an alternative derivation following the discussion of
group LASSO in \cite{tests:adaptive}. The implementation described here
was used in all simulations.
For the rest of this section let $g = \gstar$ be the index of
the group attaining the maximum on line 3 of Algorithm~\ref{algo:fs}.
%$\lambda = \norm{X_g^Ty}_2/w_g$ be the value of the maximum, and
%$r = \text{rank}(X_g)$ be the rank of the maximizing group.
For a vector $u \in \real^p$, let $u_h$ denote the coordinates
of $u$ corresponding to the columns of group $h$ in the design matrix $X$.
We can rearrange the columns of $X$ to group these adjacently, so that
\[
u^TX^T = \begin{pmatrix} u_1^T X_1^T & u_2^T X_2^T & \cdots & u_G^TX_G^T \end{pmatrix}
\]
One step of the calculation will be to find an orthonormal basis for the
linear space $\vecsp_g = \{ u \in \real^p : u_g^T X_g^T y = 0, u_h = 0
\text{ for all } h \neq g \}$ so that we can project orthogonally to
this space. If $X_g$ is a single column and $X_g^Ty \neq 0$ (which
should be the case since $g$ maximizes the absolute value of this
quantity), then the
space is trivial and the desired orthogonal projection is the identity.
Otherwise, if $X_g$ has $p_g > 1$ columns the space $\vecsp_g$
generally has dimension $p_g-1$.

\begin{equation}
 \begin{aligned}
   \label{eq:cvar}
   \sigma^2_g &= y^TX_gX_g^T H_g X_gX_g^Ty / (w_g^2 \norm{X_g^Ty}_2^2) \\
 \end{aligned}
\end{equation}
These simplify when $X_g$ is a single column, in which case $P_g = 0$,
$H_g = \Sigma$, and $\sigma^2_g = X_g^T\Sigma X_g^T/w_g^2$.
Note that $F_{\chi^2_r}$ denotes the distribution function of a $\chi^2$
random variable with $r$ degrees of freedom. This completes the
description of Algorithm~\ref{algo:pval} which we use to compute our
p-value. Additional details, including the statement and solution of
the \textsc{LinearFractional} subproblem, appear in Appendix~\ref{app:closedform}.

@@

The present work calculates p-values at each step ignoring the constraints
imposed by previous steps. Future work adjusting for all previous steps
is in progress, and may be able to give exact p-values at a step chosen
stochastically by procedures like BIC and RIC. The authors hope to
eventually release an R package implementing these methods.

>>>>

% "Title of the paper"
\title{A significance test for forward stepwise model selection}

@@

\begin{abstract}
  We apply the methods developed by \cite{significance:lasso} and
  \cite{tests:adaptive} on significance tests for penalized
  regression to forward stepwise model selection. We introduce a new
  test statistic for settings with grouped variables, allowing us to
  handle categorical variables and factor models. This
  statistic has a truncated $\chi$ distribution under the global null,
  and we apply it in forward stepwise iteratively on the residual after
  each step. The resulting method
  has the computational strengths of stepwise selection and addresses
  the problem of invalid test statistics due to model selection.
  We illustrate the flexibility of this method by applying it to
  several specialized applications of forward stepwise including a
  hierarchical interactions model and regression with polynomial splines.
\end{abstract}

@@

Consider the regression setting with a single response variable $Y$
and collection of predictor or covariate variables denoted $\mathcal X$.
One often wishes to choose a subset of variables $X \subset \mathcal X$
for modeling the response, assuming the remaining variables are
irrelevent and can be discarded without much loss of predictive
or explanatory power. Doing this in a structured and principled way
usually requires algorithmic methods for choosing $X$. One such method,
forward stepwise regression, is a procedure
that begins with an empty model and sequentially adds the best predictor
variable in each step. Because of the stochastic nature of this
algorithm---making use of the data to choose variables---the
usual $\chi^2$ and $F$-tests for significance
fail when a model has been selected this way. These tests will be
anti-conservative unless they are computed on a held-out validation
dataset. This problem is one instance of the general problem of
conducting inference and model selection using the same data, a problem
of central importance on which some recent progress has been made.

In the LASSO setting, \cite{significance:lasso} derived
a novel test statistic and its asymptotic null distribution, making
possible valid inferences after model selection using the full data.
\cite{tests:adaptive} modified and extended those results to the
group LASSO \citep{grouplasso} and other penalized regression
problems, but only under the global null hypothesis.
One of the strengths of these test statistics is that they can be
used for valid significance testing when computed on the same
data used for model selection, eliminating the need for data splitting.
This is especially important in
situations in which data splitting is not appropriate. For example,
when categorical covariates have levels with very few observations
it can be difficult or impossible to split the sample with an adequate
number of such observations occurring in each split. Furthermore, even
if sample splitting is possible, it sacrifices accuracy in the model
selection procedure and power in the subsequent inferences.

@@

As a small illustrative example of what we gain from this method,
consider a response $Y$ with ten
categorical predictors $X_g$ having between 2 and 4 levels. The true 
relationship is that $Y$ only depends on $X_1$ and $X_9$. Specifically,
if observation $i$ has covariates $X_{1,i} = j, X_{9,i} = k$ then
\[
    Y_i = \beta_{1,j} + \beta_{9,k} + \epsilon_i , \qquad
    \epsilon_i \sim N(0,1)
\]
$X_1$ has three levels with $\beta_1 = (1, .5, -1)$, $X_9$
has two levels with $\beta_9 = (.5, -.5)$. We generated a random
categorical design matrix $X$ and created an instance of this example
with $n = 40$ observations. Running forward stepwise, we calculated
both our p-value and the usual $\chi^2$ p-value at each step. In
Table~\ref{tab:example} we see that forward stepwise chooses the
truly nonzero variables first and both p-values for these are small.
However, once forward stepwise begins adding noise variables, the
$\chi^2$ p-value remains relatively small, potentially leading to 
incorrect inferences.

@@

In the next section we establish general notation and describe the
forward stepwise algorithm used throughout the paper.
Section~\ref{sec:testing} reviews some recent work on
post-selection inference
\citep{significance:lasso,tests:adaptive,lasso:fixed}
relevant to our work here, and describes a general framework
for post-selection inference based on quadratic comparisons.
Simulation results in Section~\ref{sec:simulations} show
empirically that our method performs well in settings where forward
stepwise itself performs well, and that
various stopping rules using our test statistic---including
some from \cite{sequential:fdr}---appear promising. In
Section~\ref{sec:applications} we apply
the method to several variants of forward
stepwise tailored to models with interactions and generalized additive
models, as well as to a real data example involving genomic prediction
of individual drug responses to treat HIV.

@@

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of dummy variable encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single functional pathway.
To emphasize this we use $g, h$ as covariate indices rather than the usual
$i, j$ throughout. Since single variables can be considered groups of
size 1, this includes non-grouped situations as a special
case.
%Our variant of forward stepwise uses a different method than
%usual for choosing which group to add because we are using the
%same objective in the forward stepwise procedure that was used to
%derive the test statistics in penalized regression settings.

Let $y \in \real^n$ be $n$ i.i.d. measurements of the outcome variable.
Let an integer $G \geq 2$ be the number of groups of explanatory variables.
For each $1 \leq g \leq G$ the design matrix encoding the
$g$th group is the $n \times p_g$ matrix denoted $X_g$, where $p_g$ is
the number of individual variables or columns in group $g$.
% Do we need to assume $p_g \ll n$ or $\text{rank}(X_g) = p_g$?
When a group encodes a categorical variable as indicators for the
levels of that variable, by default we use the full encoding with a
column for every level. Although this introduces collinearity, our method
does not require the design matrix to have full rank.

Denote by $p = \sum_{g=1}^Gp_g$
the total number of individual variables, so $p = G$ in the
case where all groups have size 1. Let $X$ be the matrix constructed
by column-binding the $X_g$, that is  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}
X = \begin{pmatrix} X_1 & X_2 & \cdots & X_G  \end{pmatrix}
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We also allow weights $w_g$ for each group. These
weights act like penalties or costs, so increasing $w_g$ makes it
more difficult for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily for calibration purposes,
but throughout we set them all constant (equal to 1) and normalize
groups by the Frobenius norm of their corresponding submatrices.

With each group we associate the $p_g \times 1$ coefficient vector
$\beta_g$, and write $\beta$ for the $p \times 1$ vector constructed
by stacking all of the $\beta_g$ in order.  Finally, our model for the
response is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gmodel}
y & = X \beta + \sigma \epsilon \\
   & = \sum_{g=1}^G X_g \beta_g + \sigma \epsilon
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\epsilon$ is noise. We assume Gaussian noise
$\epsilon | X \sim N(0, \Sigma)$ with known covariance matrix $\Sigma$.

@@

\subsection{Description of the forward stepwise algorithm}

First, the user must specify the maximum number of steps allowed,
which we denote \textit{steps}.
%(consistent with the \textit{R} function for stepwise).
To enable our test statistic computations,
\textit{steps} should be at most $\min (n, G) - 1$, but it is
computationally desirable to set it as low as possible while
safely larger than the sparsity of $\beta$. 
Then forward stepwise may recover all the nonzero
coefficients of $\beta$ and terminate without performing much
additional computation.
Of course the sparsity is usually unknown, so this requires guesswork.
In our implementation we treat the active set $A$ as an ordered list to easily track the order of groups entering the model.

@@

To describe the previous work we require some facts about the solution path
of the LASSO.
Recall the LASSO estimator is given by
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:lasso}
\displaystyle \hat \beta(\lambda) = \argmin_{\beta \in \real^p} \frac{1}{2} \| y - X \beta \|_2^2 +
   \lambda \| \beta \|_1 \\
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The following facts are summarized in \cite{significance:lasso,tibshirani_LASSO_uniqueness}.

@@

Assuming groups of size one and $\Sigma = \sigma^2 I$, the covariance test is given at the first step by
\begin{equation}
\label{eq:covtest}
T_1 = \frac{\lambda_1(\lambda_1-\lambda_2)}{\sigma^2} \overset{n,p \to \infty}{\to} \text{Exp}(1).
\end{equation}
This is a hypothesis test for including the first variable in the LASSO
path, with large values of the test statistic being evidence against the
global null.
In \cite{tests:adaptive} it is pointed out 
\begin{equation}
\label{test:exact}
\exp\left(- \frac{\lambda_1(\lambda_1-\lambda_2)}{\sigma^2} \right) \approx \frac{1 - \Phi(\lambda_1/\sigma)}{1 - \Phi(\lambda_2 / \sigma)} \overset{D}{=} \text{Unif}(0,1).
\end{equation}
The right hand side has an exact, finite sample null distribution.
Further, this new statistic can be understood as the survival function of $\lambda_1 = \|X^Ty\|_{\infty}$
{\em conditional} on which variable achieves $\lambda_1$ as well as its sign.
The limiting results about later steps in \cite{significance:lasso} can
%(formally)
be interpreted as
recursively applying $T_1$ to the variables that had not previously been selected by LARS. In forward stepwise with groups of size 1, we recursively apply the right hand side test statistic in the same manner. With larger groups, we use an analogous test statistic to the right hand side of \eqref{test:exact}. The test statistic is presented as an example in \cite{tests:adaptive} though we re-derive it here in simpler form. The test draws inspiration
from the group LASSO.

The \emph{group LASSO estimator}, a generalization of the LASSO,
 is a solution to the following penalized
least-squares convex problem
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gsoln}
\displaystyle \hat \beta_\lambda = \argmin_{\beta \in \real^p} \frac{1}{2} \| y - X \beta \|_2^2 +
   \lambda {\cal P}(\beta) \\
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
with the group penalty
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
  \begin{aligned}
  \label{eq:gpen}
    {\cal P}(\beta)&= \sum_{g=1}^G w_g \| \beta_g \|_2 \\
  \end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The parameter $\lambda \geq 0$ enforces sparsity in groups: for large
$\lambda$ most of the $\beta_g$ will be zero vectors. The weights
$w_g$ are usually taken to be $\sqrt {p_g}$ to normalize the penalty
across groups with different sizes.  This can also be accomplished
by scaling the corresponding submatrices by their Frobenius norms
and setting all weights equal.
Note that this includes the usual LASSO estimator as a
special case when all groups are of size 1, since then the
penalty term is the $\ell_1$-norm of $\beta$. The solution path of the group
LASSO has similar properties to the LASSO case, however it is not
generally piecewise linear.

@@

With non-trivial groups, however, the event that $g^*(y)$, the first
group chosen, is equal to some fixed group $g$ is generally not given by
a set of affine constraints. Rather, considering line 3 of Algorithm~\ref{algo:fs}, we see
\begin{equation}
\label{eq:first:quadratic}
\begin{aligned}
\{g^*(y)=g\} &= \{ \|X_g^Ty\|_2 /w_g \geq \|X_h^Ty\|_2 /w_h , \ \forall h \neq g\} \\
&=  \{ y^T(X_g^TX_g)y / w_g^2 - y^T(X_h^TX_h)y / w_h^2 \geq 0, \ \forall h \neq g\} \\
\end{aligned}
\end{equation}
Hence, our selection event can be expressed as 
the intersection of a list of quadratic inequalities. This non-affine selection event is stylized in Figure
\ref{fig:curved}.
In the rest of this section, we consider an arbitrary selection procedure
$S$ that returns one of a set of possible outcomes $s \in {\cal S}$ by a set of quadratic inequalities.
That is, for all possible outcomes $s \in {\cal S}$ there is a set of indices
$I(s)$ such that
\begin{equation}
\label{eq:selection}
\begin{aligned}
\{S(y)=s\} = \cap_{i \in I(s)} \{y: y^TQ_iy + a_i^Ty\leq b_i \}.
\end{aligned}
\end{equation}
The quadratic forms are not assumed non-negative definite, but we can, without loss of generality
assume they are symmetric.
As in \cite{lasso:fixed} we will be interested in some randomly chosen contrast
$\eta(S(y))^T\mu$.
In our grouped selection procedure, the selection rule is $S(y)=g^*(y)$
and $\eta$ is typically 
$$
\eta(S(y)) = \eta(g^*(y)) = \frac{X_{g^*(y)}^Ty}{\|X_{g^*(y)}^Ty\|_2}.
$$
This choice of $\eta$ reflects our goal of calculating a p-value for inclusion of the variable $g^*$ chosen by forward stepwise.

Having fixed $\eta$, we proceed to study this contrast by slicing through the selection event along a ray with direction $\eta$ that passes through $y$. 
That is, we need to find
$$
\begin{aligned}
\lefteqn{
\left\{t \in \mathbb R : S(y+t \cdot \eta)=s \right\} } \\
 & \qquad = \cap_{i \in I(s)} \left\{t \in \mathbb R : (y+t \cdot \eta)^T Q_i (y+t \cdot \eta) + a_i^T(y+t \cdot \eta) \leq b_i\right\}. \\
\end{aligned}
$$
\begin{figure}
\begin{center}
\resizebox{!}{3in}{\input{../figs/curved.pdf_edit_tex}}
\end{center}
\caption{ \em
A generic selection event, given by the intersection of a list
of quadratic inequalities. Note that the event need not be simply connected.
}
\label{fig:curved}
\end{figure}
The slice for any particular inequality is
\begin{equation}
\begin{aligned}
\label{eq:quadslice}
\lefteqn{
\left\{t \in \mathbb R : (y+t\cdot \eta)^T Q_i (y+t \cdot \eta) + a_i^T(y+t \cdot \eta) \leq b_i\right\}} \\
 & \qquad = \left\{t \in \mathbb R : t^2 \cdot (\eta^TQ_i\eta) + t \cdot(2 y^TQ_i\eta + a_i^T\eta) + y^TQ_iy + a_i^Ty - b_i \leq 0 \right\} \\
\end{aligned}
\end{equation}
This can be computed explicitly, and results in either the empty set, a single interval (possibly infinite) or the union of 
two disjoint infinite intervals. The intersection over all $I(s)$ is therefore also computable, yielding a form for the slice
\begin{equation}
\label{eq:slice}
\{g^*(y)=g, \eta(g^*(y))=\eta\}.
\end{equation}
In Figure \ref{fig:curved}, for every $\eta$, the slice \eqref{eq:slice} is a function
of $P_{\eta}^{\perp}y$. This amounts to a proof of the following lemma.

@@

{\bf Proof:}
For any fixed $s$, decompose $y$ as
$$
(Z^{\perp}_s,Z_s) = (P_s^{\perp}y, X_s^Ty).
$$
Under $H_{0,s}:X_s^T\mu=0$, the density of $Z_s$ can be written as
$$
(2 \pi \Sigma_s \sigma^2)^{-k(s)/2} \exp \left(-\frac{1}{2 \sigma^2} z^T\Sigma_s^{-1}z \right)
$$
with $\Sigma_s = X_s^TX_s$.
Transforming to polar coordinates $(R,U)$ yields a density
$$
(2 \pi \Sigma_s \sigma^2)^{-k(s)/2} r^{k(s)-1} \exp \left(-\frac{r^2}{2 \sigma^2} u^T\Sigma_s^{-1}u \right), \qquad r \geq 0, \|u\|_2=1.
$$
Conditioning on $U$ shows that $R=\|X_s^Ty\|_2|U$ has distribution proportional to a $\chi_{k(s)}$. Finally, observe that 
the scale of the $\chi$ above is given by $\theta_s$. \qed

Lemma \ref{eq:test:dbn} gives an explicit characterization of our p-value
that can also be used to calculate it in practice. An implementation only needs to compute the intersection of all the intervals implied by each quadratic constraint \eqref{eq:quadslice} and then evaluate the $\chi$ distribution truncated to the intersection of these intervals at the observed value.
For an alternative
derivation including a detailed algorithm description for the implementation used throughout this paper see Appendix~\ref{app:algo}.

After adding the first group in forward stepwise and computing its p-value, we orthogonalize the response $y$ and all remaining groups. This imposes some constraints on $y$ that in principle should be used when computing p-values at subsequent steps. For now we ignore these constraints and iterate the global null test, but work on incorporating all the known constraints is ongoing. We believe that the effect of not tracking all constraints is what causes the nominal p-value to be increasingly stochastically larger than uniform further down the forward stepwise path, and only truly uniform at the first step where all remaining variables are noise.

%$\sigma \|X_s\eta_s\|_2$.

@@

@@

The present work calculates p-values at each step ignoring the constraints
imposed by previous steps. Future work adjusting for all previous steps
is in progress, and may be able to give exact p-values at a step chosen
stochastically by procedures like BIC and RIC. The authors will
release an R package implementing these methods.
